---
title: "my_markdown4"
author: "aisha"
date: "2025-12-09"
output: html_document
---
# 1. Introduction
This report demonstrates a complete analysis workflow using R markdown. It includes data loading, cleaning, analysis, visualization and interpretation.

#2 load libraries
```{r, message=FALSE}
library(tidyverse)
library(knitr)
```
# 3 import Data
```{r}
data <- tibble(
  id = 1:10,
  age = c(23, 45, 34, 29, 50,41, 38, 27, 32, 46),
  score = c( 55, 72, 68, 60, 80, 75, 70, 58, 65, 78)
)
data
```
# 4 Data summary
## 4.1 summary statistics
```{r}
summary(data)
```
## 4.2 Table of Mean Values
```{r}
mean_table <- data %>%
  summarise(
    mean_age = mean(age),
    mean_score = mean(score)
  )
kable(mean_table)
```
# 5 Visualization
## 5.1 scatter plot
```{r}
ggplot(data, aes(x = age, y = score))+
  geom_point(size = 3)+
  geom_smooth(method = ,
              
              se = FALSE)
```
## 5.2 Histogram
```{r}
ggplot(data, aes( x = score))+
  geom_histogram(bins = 5)
```
# 6 simple statistical Analysis
## 6.1 Linear Regression
```{r}
model <- lm (score ~ age, data = data)
summary(model)
```
## 6.2 Interpretation
The regression model shows how **age predicts score**.
A positive coefficient means older participants tend to score higher.
check the *p-value* to see if the relationship is statistically significant.
# 7 . conclusion
This report shows a full workflow:
-Data import
-summary statistics
-Visualization
-Regression analysis
- Interpretation









