---
title: "Writing for loops"
author: "hirbo"
date: "2026-03-16"
output: html_document
---
#Writing for loops
```{r}
library(dplyr)
```
#simulate some data
```{r}
set.seed(123)
df_xyz <- tibble(
  x = rnorm(10),
  y = rnorm(10),
  z = rnorm(10)
) %>% 
  print()
```
#to calculate the mean without repetition using across function
```{r}
df_xyz %>%
  summarise(
    across(
      .cols  = everything(),
      .fns   = mean,
      .names = "{col}_mean"
    )
  )
```
#using a for loop 
```{r}
xyz_means <- vector("double", ncol(df_xyz))

for(i in seq_along(df_xyz)) {
  xyz_means[[i]] <- mean(df_xyz[[i]])
}
```
```{r}
xyz_means
```
#using the dolar sign
```{r}
mean(df_xyz$x)
mean(df_xyz$y)
mean(df_xyz$z)
```
#the for() function
```{r}
for (i in 1) { print(i)
  
}
```
#for loop to iterate over a data frame
```{r}
for(i in df_xyz) {
  print(i)
}
```
```{r}
xyz_means <- df_xyz %>%
  summarise(
    across(
      .cols  = everything(),
      .fns   = mean,
      .names = "{col}_mean"
    )
  )
```
```{r}
xyz_means <- for(i in df_xyz) {
  mean(i)
}
```
#create object that can store data for loop
```{r}
my_vec <- vector()
my_vec
```
```{r}
xyz_means <- vector("numeric", 3)
xyz_means
```
#update xyz means
```{r}
for(i in df_xyz) {
  xyz_means <- mean(i)
}
xyz_means
```
```{r}
df_xyz <- df_xyz %>% 
  mutate(a = rnorm(10)) %>% 
  print()
```
```{r}
xyz_means <- vector("numeric", length(df_xyz)) # Using length() instead of 3

for(i in seq_along(df_xyz)) { # Using seq_along() instead of c(1, 2, 3)
  xyz_means[[i]] <- mean(df_xyz[[i]])
}

xyz_means
```
#writing function our for loop
```{r}
multi_means <- function(data) {
  # Create a structure to contain results
  result <- vector("numeric", length(data))
  
  # Iterate over each column of data
  for(i in seq_along(data)) {
    result[[i]] <- mean(data[[i]])
  }
  
  # Return the result
  result
}
```
```{r}
multi_means(df_xyz)
```
#because we’ve made the for loop code inside of the function body flexible with length() and seq_along() we can easily pass any other data frame (with all numeric columns) to our function like this:
```{r}
set.seed(123)
new_df <- tibble(
  age    = rnorm(10, 50, 10),
  height = rnorm(10, 65, 5),
  weight = rnorm(10, 165, 10)
) %>% 
  print()
```
#for loop to return the result with informative names
```{r}
xyz_means <- vector("numeric", length(df_xyz))

for(i in seq_along(df_xyz)) {
  xyz_means[[i]] <- mean(df_xyz[[i]])
  names(xyz_means)[[i]] <- paste0(names(df_xyz)[[i]], "_mean") # Name results here
}

xyz_means
```
#update the multi_means function
```{r}
multi_means <- function(data) {
  # Create a structure to contain results
  result <- vector("numeric", length(data))
  
  # Iterate over each column of data
  for(i in seq_along(data)) {
    result[[i]] <- mean(data[[i]])
    names(result)[[i]] <- paste0(names(data)[[i]], "_mean")
  }
  
  # Return the result as a tibble
  as_tibble(as.list(result))
}
```
```{r}
multi_means(new_df)
```
#using for loops for data transfer
```{r}
chooseCRANmirror()
```
```{r}
install.packages("readxl")
```
```{r}
library(readxl)
```
```{r}
houston <- read_excel(
  "city_ses.xlsx",
  sheet = "Houston"
) %>% 
  print()
```
#loops for data management
```{r}
people_1 <- tribble(
  ~id_1, ~name_first_1, ~name_last_1, ~street_1,
  1,     "Easton",      NA,           "Alameda",
  2,     "Elias",       "Salazar",    "Crissy Field",
  3,     "Colton",      "Fox",        "San Bruno",
  4,     "Cameron",     "Warren",     "Nottingham",
  5,     "Carson",      "Mills",      "Jersey",
  6,     "Addison",     "Meyer",      "Tingley",
  7,     "Aubrey",      "Rice",       "Buena Vista",
  8,     "Ellie",       "Schmidt",    "Division",
  9,     "Robert",      "Garza",      "Red Rock",
  10,    "Stella",      "Daniels",    "Holland"
)
```
```{r}
people_2 <- tribble(
  ~id_2, ~name_first_2, ~name_last_2, ~street_2,
  1,     "Easton",      "Stone",      "Alameda",
  2,     "Elas",        "Salazar",    "Field",
  3,     NA,            "Fox",        NA,
  4,     "Cameron",     "Waren",      "Notingham",
  5,     "Carsen",      "Mills",      "Jersey",
  6,     "Adison",      NA,           NA,
  7,     "Aubrey",      "Rice",       "Buena Vista",
  8,     NA,            "Schmidt",    "Division",
  9,     "Bob",         "Garza",      "Red Rock",
  10,    "Stella",      NA,           "Holland"
)
```
```{r}
people <- people_1 %>% 
  bind_cols(people_2) %>% 
  print()
```
```{r}
is_match <- function(value_1, value_2) {
  result <- value_1 == value_2
  result <- if_else(is.na(result), FALSE, result)
  result
}
```
```{r}
people %>% 
  mutate(
    name_first_match = is_match(name_first_1, name_first_2),
    name_last_match  = is_match(name_last_1, name_last_2),
    street_match     = is_match(street_1, street_2)
  ) %>% 
  # Order like columns next to each other for easier comparison
  select(id_1, starts_with("name_f"), starts_with("name_l"), starts_with("s"))
```
#remove unnecessary repetition using loops
```{r}
cols <- c("name_first", "name_last", "street")

for(i in seq_along(cols)) {
  col_1   <- paste0(cols[[i]], "_1")
  col_2   <- paste0(cols[[i]], "_2")
  new_col <- paste0(cols[[i]], "_match")
  people[[new_col]] <- is_match(people[[col_1]], people[[col_2]])
}
```
```{r}
people %>% 
  select(id_1, starts_with("name_f"), starts_with("name_l"), starts_with("s"))
```
```{r}
cols <- c("name_first", "name_last", "street")

for(i in seq_along(cols)) {
  col_1 <- paste0(cols[[i]], "_1")
  col_2 <- paste0(cols[[i]], "_2")
  print(people[[col_1]])
}
```
#pass each vector to our is_match function
```{r}
cols <- c("name_first", "name_last", "street")

for(i in seq_along(cols)) {
  col_1 <- paste0(cols[[i]], "_1")
  col_2 <- paste0(cols[[i]], "_2")
  print(is_match(people[[col_1]], people[[col_2]]))
}
```
```{r}
cols <- c("name_first", "name_last", "street")

for(i in seq_along(cols)) {
  col_1   <- paste0(cols[[i]], "_1")
  col_2   <- paste0(cols[[i]], "_2")
  new_col <- paste0(cols[[i]], "_match")
  people[[new_col]] <- is_match(people[[col_1]], people[[col_2]])
}
```
```{r}
people %>% 
  select(id_1, starts_with("name_f"), starts_with("name_l"), starts_with("s"))
```
#for loops for Analysis
#simulate some data
```{r}
study <- tibble(
  age       = c(32, 30, 32, 29, 24, 38, 25, 24, 48, 29, 22, 29, 24, 28, 24, 25, 
                25, 22, 25, 24, 25, 24, 23, 24, 31, 24, 29, 24, 22, 23, 26, 23, 
                24, 25, 24, 33, 27, 25, 26, 26, 26, 26, 26, 27, 24, 43, 25, 24, 
                27, 28, 29, 24, 26, 28, 25, 24, 26, 24, 26, 31, 24, 26, 31, 34, 
                26, 25, 27, NA),
  age_group = c(2, 2, 2, 1, 1, 2, 1, 1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 
                1, 1, 1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 1, 1, 1, 1, 1, 1, 
                1, 1, 1, 2, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 2, 1, 1, 2, 
                2, 1, 1, 1, NA),
  gender    = c(2, 1, 1, 2, 1, 1, 1, 2, 2, 2, 1, 1, 2, 1, 1, 1, 1, 2, 2, 1, 1, 
                1, 1, 2, 1, 1, 2, 1, 1, 1, 2, 1, 1, 2, 2, 1, 2, 2, 1, 2, 2, 1, 
                1, 1, 1, 1, 1, 1, 1, 2, 2, 1, 1, 1, 1, 2, 2, 1, 1, 2, 1, 2, 1, 
                1, 1, 2, 1, NA),
  ht_in     = c(70, 63, 62, 67, 67, 58, 64, 69, 65, 68, 63, 68, 69, 66, 67, 65, 
                64, 75, 67, 63, 60, 67, 64, 73, 62, 69, 67, 62, 68, 66, 66, 62, 
                64, 68, NA, 68, 70, 68, 68, 66, 71, 61, 62, 64, 64, 63, 67, 66, 
                69, 76, NA, 63, 64, 65, 65, 71, 66, 65, 65, 71, 64, 71, 60, 62, 
                61, 69, 66, NA),
  wt_lbs    = c(216, 106, 145, 195, 143, 125, 138, 140, 158, 167, 145, 297, 146, 
                125, 111, 125, 130, 182, 170, 121, 98, 150, 132, 250, 137, 124, 
                186, 148, 134, 155, 122, 142, 110, 132, 188, 176, 188, 166, 136, 
                147, 178, 125, 102, 140, 139, 60, 147, 147, 141, 232, 186, 212, 
                110, 110, 115, 154, 140, 150, 130, NA, 171, 156, 92, 122, 102, 
                163, 141, NA),
  bmi       = c(30.99, 18.78, 26.52, 30.54, 22.39, 26.12, 23.69, 20.67, 26.29, 
                25.39, 25.68, 45.15, 21.56, 20.17, 17.38, 20.8, 22.31, 22.75, 
                26.62, 21.43, 19.14, 23.49, 22.66, 32.98, 25.05, 18.31, 29.13, 
                27.07, 20.37, 25.01, 19.69, 25.97, 18.88, 20.07, NA, 26.76, 
                26.97, 25.24, 20.68, 23.72, 24.82, 23.62, 18.65, 24.03, 23.86, 
                10.63, 23.02, 23.72, 20.82, 28.24, NA, 37.55, 18.88, 18.3, 
                19.13, 21.48, 22.59, 24.96, 21.63, NA, 29.35, 21.76, 17.97, 
                22.31, 19.27, 24.07, 22.76, NA),
  bmi_3cat  = c(3, 1, 2, 3, 1, 2, 1, 1, 2, 2, 2, 3, 1, 1, 1, 1, 1, 1, 2, 1, 1, 
                1, 1, 3, 2, 1, 2, 2, 1, 2, 1, 2, 1, 1, NA, 2, 2, 2, 1, 1, 1, 1, 
                1, 1, 1, 1, 1, 1, 1, 2, NA, 3, 1, 1, 1, 1, 1, 1, 1, NA, 2, 1, 
                1, 1, 1, 1, 1, NA)
) %>% 
  mutate(
    age_group = factor(age_group, labels = c("Younger than 30", "30 and Older")),
    gender    = factor(gender, labels = c("Female", "Male")),
    bmi_3cat  = factor(bmi_3cat, labels = c("Normal", "Overweight", "Obese"))
  ) %>% 
  print()
```
# use across() with pivot_longer() to remove repetition and get our results into a format that were easier to read an interpret
```{r}
summary_stats <- study %>% 
  summarise(
    across(
      .cols  = c(age, ht_in, wt_lbs, bmi),
      .fns   = list(
        n_miss = ~ sum(is.na(.x)),
        mean   = ~ mean(.x, na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        min    = ~ min(.x, na.rm = TRUE),
        max    = ~ max(.x, na.rm = TRUE)
      ),
      .names = "{col}-{fn}" # This is the new part of the code
    ) 
  ) %>% 
  print()
```
```{r}
study %>% 
  summarise(
    across(
      .cols = c(age_group, gender, bmi_3cat),
      .fns  = count
    )
  )
```
```{r}
study %>% 
  summarise(
    across(
      .cols = c(age, ht_in),
      .fns  = ~ mean(.x, na.rm = TRUE)
    )
  )
```
```{r}
# Structure 1. An object to contain the results.
  # Create the data frame structure that will contain our results
cat_table <- tibble(
  variable = vector("character"), 
  category = vector("character"), 
  n        = vector("numeric")
) 

# Structure 2. The actual for loop.
  # For each column, get the column name, category names, and count.
  # Then, add them to the bottom of the results data frame we created above.
for(i in c("age_group", "gender", "bmi_3cat")) {
  cat_stats <- study %>% 
    count(.data[[i]]) %>% # Use .data to refer to the current data frame.
    mutate(variable = names(.)[1]) %>% # Use . to refer to the result to this point.
    rename(category = 1)
  
  # Here is where we update cat_table with the results for each column
  cat_table <- bind_rows(cat_table, cat_stats)
}
```
```{r}
cat_table
```
#create result data frame structure
```{r}
cat_table <- tibble(
  variable = vector("character"), 
  category = vector("character"), 
  n        = vector("numeric")
) 
```
```{r}
str(cat_table)
```
#using for loops
```{r}
for(i in c("age_group", "gender", "bmi_3cat")) {
  cat_stats <- study %>% 
    count(.data[[i]]) %>%
    mutate(variable = names(.)[1]) %>%
    rename(category = 1)
  
  cat_table <- bind_rows(cat_table, cat_stats)
}
```
```{r}
cat_stats <- study %>% 
  count(.data[[i]]) %>%
  mutate(variable = names(.)[1]) %>%
  rename(category = 1)
```
```{r}
cat_stats <- study %>% 
  count(age_group) %>%
  mutate(variable = names(.)[1]) %>%
  rename(category = 1) %>% 
  print()
```
#analysis one variable at a time
```{r}
cat_stats <- study %>% 
  count(age_group) %>% 
  mutate(variable = names(.)[1]) %>% 
  rename(category = age_group) %>% 
  print()
```
#create empty result table
```{r}
cat_table <- tibble(
  variable = vector("character"), 
  category = vector("character"), 
  n        = vector("numeric")
) %>%
  print()
```
#dtaframe results for agegroup
```{r}
cat_stats <- study %>% 
  count(age_group) %>%
  mutate(variable = names(.)[1]) %>% 
  rename(category = 1) %>% 
  print()
```
```{r}
for(i in c(age_group, gender, bmi_3cat)) {
  study %>% 
    count(i) %>% 
    mutate(variable = names(.)[1]) %>% 
    rename(category = 1)
}
```
```{r}
for(i in c("age_group", "gender", "bmi_3cat")) {
  study %>% 
    count(i) %>% 
    mutate(variable = names(.)[1]) %>% 
    rename(category = 1)
}
```
```{r}
for(i in c("age_group", "gender", "bmi_3cat")) {
  study %>% 
    count(.data[[i]]) %>% 
    mutate(variable = names(.)[1]) %>% 
    rename(category = 1) %>% 
    print()
}
```
```{r}
# Create the data frame structure that will contain our results
cat_table <- tibble(
  variable = vector("character"), 
  category = vector("character"), 
  n        = vector("numeric")
) 

# For each column, get the column name, category names, and count.
# Then, add them to the bottom of the results data frame we created above.
for(i in c("age_group", "gender", "bmi_3cat")) {
  cat_stats <- study %>% 
    count(.data[[i]]) %>% # Use .data to refer to the current data frame.
    mutate(variable = names(.)[1]) %>% # Use . to refer to the current data frame.
    rename(category = 1)
  
  # Here is where we update cat_table with the results for each column
  cat_table <- bind_rows(cat_table, cat_stats)
}
cat_table
```
```{r}
cat_table %>% 
  group_by(variable) %>% 
  mutate(
    percent = n / sum(n) * 100
  )
```
#write our own function
```{r}
cat_stats <- function(data, ...) {
  # Create the data frame structure that will contain our results
  cat_table <- tibble(
    variable = vector("character"), 
    category = vector("character"), 
    n        = vector("numeric")
  ) 
  
  # For each column in ..., get the column name, category names, and count.
  # Then, add them to the bottom of the results data frame we created above.
  for(i in c(...)) {
    stats <- data %>% 
      count(.data[[i]]) %>% # Use .data to refer to the current data frame.
      mutate(variable = names(.)[1]) %>% # Use . to refer to the current data frame.
      rename(category = 1)
    
    # Here is where we update cat_table with the results for each column
    cat_table <- bind_rows(cat_table, stats)
  }
  # Return results
  cat_table
}
```
```{r}
cat_stats(study, "age_group", "gender", "bmi_3cat")
```


